Master's Project Journal

### 2019-04-20 thesis/Background.tex
79git statusgit statusgit statusgit statusgit statusgit statusgit statusgit statusgit statusgit statusgit statusgit status

In this design the oscillators are once again laid out in a Cartesian grid, with each node coupled to their neighbours in phase. As this is now a digital system the coupling is carried out using digital phase comparators, which attempt to measure the phase difference between two oscillators. Figure \ref{fig:eldar_node} shows high level detail of the architecture of both the entire clocking system and that of an individual node in the design. The digital nature of this architecture brings with it a number of advantages over traditional analogue implementations, as it can benefit from advancements in digital circuit design suites, be reconfigurable and programmable and has a significantly greater immunity to perturbations inherent to its digital nature, as the exact voltage of signals is of no importance \cite{zianbetov2013phd}. This last advantage is of particular use in a digital environment, as otherwise there is potential for clock degradation resulting from switching of transistors. The drawback of the switch to a digital architecture however is the presence of quantisation. Analogue designs both deliver continuous control signals to the oscillators and have a continuous phase detection capability, unlike a digital system where these actions are carried out with fixed resolution.
\begin{figure}[!h]
	\centering
	\includegraphics[scale=0.4]{../ccirc_2013_arch}
	\caption[Architecture of the ADPLL network and of a single node]{Architecture of the ADPLL network and of a single node \cite{zianbetov2013distributed}.}
	\label{fig:eldar_node}
\end{figure}

Looking at the design of a given node it is notable that the function carried out by the Error Combiner is akin to a average, therefore as mentioned by Pratt and Nyugen, there is potential to a mode locked equilibrium in which the oscillators are not synchronised. In their paper, they presented a method where initial start-up was performed uni-directionally and, once all nodes are close to alignment, full connectivity could be restored, however this was not viable in an analogue system as reconfigurability was not an option \cite{pratt1995distributed}. Javidan \textit{et al} (2011) found that this was in fact the case, and stated:
\begin{quote}
	[Mode-locking] is solved in a simple and original way, by a dynamic reconfiguration of the network interconnection topology at the starting stage.
\end{quote}
In creating an entirely digital system, Zianbetov and Shan could easily reconfigure the network topology and by implement a uni-directional start-up avoid the problem of convergence into a mode locked state, without having to design a non-linear phase detector.

\section{\acl{ADPLL} Architecture}
As indicated in Figure \ref{fig:mulkeen_pll} the three main building blocks of a conventional \ac{PLL} are the \ac{PD}, \ac{LF} and \ac{VCO}. In an \ac{ADPLL} these blocks are then replaced by their digital counterparts, necessitating quantisation in order to remain physically realisable. The ``All-Digital'' moniker is a misnomer as the oscillator and \acl{PD} are usually both implemented by mixed signal circuits.
\begin{figure}[h]
	\centering
	\includegraphics[scale=0.5]{../tex_files/mulkeen_pll}
	\caption[Block fiagram of a \acl{PLL}]{Block fiagram of a \acl{PLL}, \textit{Wireless Systems Notes}, B. Mulkeen (2017).}
	\label{fig:mulkeen_pll}
\end{figure}

\subsection{Digitally Controlled Oscillator}
In a digital system there are a very limited number of voltages representable, most commonly just two, so using a voltage to control the oscillator is not a viable strategy. Instead a fixed bit width signal is used to control the oscillator's period, selecting the number of inverters in a ring oscillator or the varactor configuration of a travelling wave oscillator \cite{chen2011rotary}. The decisions made in the design of the \ac{DCO}, or \ac{NCO}, determine many of the other \ac{ADPLL} parameters. While tuning range and centre frequency, as well as linearity, carry over from the analogue counterpart, a \ac{DCO} also has a frequency step which in combination with the bit width of the control signal determines the range over which the oscillator can be tuned. Figure \ref{fig:my_ring} illustrates a basic ring oscillator design. A ring oscillator is an inherently unstable circuit composed of an odd number of inverters connected in a circle, which allows a signal to propagate infinitely, with the signal at any point in the circuit appearing as a square wave. The half-period of this oscillator is the time taken for the signal to propagate once through the chain, $n$ times the propagation delay through one inverter. The frequency of operation can then be set by modulating the length of the chain, in steps of two inverters to maintain an odd number, by means of \texttt{f\_select}. The main impact of output frequency quantisation is that only frequencies which are integer multiples of the frequency step away from the centre frequency can be easily reproduced, with intermediate values only obtainable in a manner akin to Fractional-N synthesis with the control code toggling back and forth. This acts as a source of jitter in the system.
\begin{figure}[h]
	\centering
	\includegraphics[scale=0.275]{../inverter_chain}
	\caption[Basic Ring/Inverter Chain Oscillator]{Basic ring/inverter chain oscillator.}
	\label{fig:my_ring}
\end{figure}

@@

\begin{figure}[h]
	\centering
	\includegraphics[scale=0.35]{../shan_bb_pd}
	\caption[Bang-bang phase/frequency detector architecture]{Bang-bang phase/frequency detector architecture \cite{shan2014phd}.}
	\label{fig:shan_bb_pd}
\end{figure}

\subsection{Digital Loop Filter}
\begin{figure}[h]
	\centering
	\includegraphics[width=0.6\textwidth]{../simple_pi}
	\caption[Basic \ac{PI} controller architecture]{Basic \ac{PI} controller architecture.}
	\label{fig:my_simple_pi}
\end{figure}%TODO ask mulkeen if this is IIR
The \acl{LF} in an \ac{ADPLL} can be implemented as a \ac{PI} controller, as only a low-pass filter is required, such as that in Figure \ref{fig:my_simple_pi}. In the case of a node in an \ac{ADPLL} network the input of this filter is a weighted of the phase difference relative to the neighbouring local clocking areas. In one example topology a digital system the proportional section can be implemented by a simple multiplier, whereas the integral path is constructed by adding the result of a multiplication by the proportional gain to an accumulator. This delayed summation can be easily implemented by an accumulator to which the current value of the multiplication is added each cycle, and as such the system has an infinite impulse response. The value of these gains determine the response and stability of the \ac{ADPLL} network. The transfer function of such a controller is given by \cite{shan2014phd}:
\begin{equation*}
	H(z) = \alpha + \beta\frac{1}{1-z^{-1}} = \frac{(\alpha + \beta) - \alpha z^{-1}}{1-z^{-1}}
\end{equation*}
It has been found by Koskin \textit{et al} (2018) that stable operation can be achieved when the integral gain, $k_i$, is less than half the proportional gain, $k_p$ \cite{koskin2018generation}. In the same study a range of values was found which would produce low jitter operation of the network. As these values are all less than one, the filter must implement fixed point arithmetic in an effort to maintain the simplicity of the clock distribution network, rather than incurring the complexity penalty of floating point calculations.

@@

@@

\ac{FPGA}s have been used in other fields to simulate and experiment with new technologies, although not all of these attempt to implement the mixed-signal circuitry using primitive elements. Fernandez-Alvarez \textit{et al} (2016) proposed a method suited to higher end \ac{FPGA}s in which a co-processor on the \ac{FPGA} simulates the mixed-signal circuitry while the digital section of the design is implemented on the \ac{FPGA} itself \cite{fernandez2017hw}. While the interfacing between hardware and software remains a challenge they found:
\begin{quotation}
	Obtained data are compared to the data obtained by means of using PSIM and ModelSim co-simulation. The proposed solution speeds up the evaluation in around one order of magnitude keeping the accuracy. The output signal differs in less than 0.6 mV (RMSD).
\end{quotation}

Mixed-signal circuits were, however, simulated in hardware by \'{O}scar Luc\'{i}a \textit{et al} (2011) on an \ac{FPGA} in order to overcome the excessive time penalty imposed by software based simulations that required the behaviour of both a digital and mixed-signal peripheral and that of a micro controller running code in C to be simulated side by side \cite{lucia2011real}. They concluded
\begin{quotation}
	... the proposed system provides a versatile and fast method to develop ad hoc control architectures, avoiding the need for time-consuming mixed-signal simulations and the risk of damaging the actual power converter implementation.

\end{quotation}
By carrying out simulations on an \ac{FPGA}, Guanhua Wang \textit{et al} (2013) achieved a 3000 times decrease in run-time when compared to an identical simulation in MATLAB used for the verification of a calibration algorithm for successive approximation analogue-to-digital converters \cite{wang2013fast}. Many other examples exist of \ac{FPGA}s used for the simulation of mixed-signal or \acl{RF} circuits in literature.
%TODO Elena had suggestions?

@@

Both above forms of measurement assume a large number of sequential measurements, however, there are other ways that jitter and skew can be calculated \cite{AN10007}. Phase jitter is an important characteristic for communications systems as it can be used to calculate phase noise, important to ensure spurious emissions are within regulations. Long term, or accumulated, jitter represents the cumulative effect of jitter on the signal over several cycles. Long term jitter in particular affects \acs{RADAR} as it will manifest itself as a Doppler shift in the return signal. For an \ac{PLL} network, with an appropriately designed filter, the long term jitter should be zero so long as the reference signal is stable. %TODO ask Brian about this one

>>>>

In this design the oscillators are once again laid out in a Cartesian grid, with each node coupled to their neighbours in phase. As this is now a digital system the coupling is carried out using digital phase comparators, which attempt to measure the phase difference between two oscillators. Figure \ref{fig:eldar_node} shows high level detail of the architecture of both the entire clocking system and that of an individual node in the design. The digital nature of this architecture brings with it a number of advantages over traditional analogue implementations, as it can benefit from advancements in digital circuit design suites, be reconfigurable and programmable and has a significantly greater immunity to perturbations inherent to its digital nature, as the exact voltage of signals is of no importance \cite{zianbetov2013phd}. This last advantage is of particular use in a digital environment, as otherwise there is potential for clock degradation resulting from switching of transistors. The drawback of the switch to a digital architecture however is the presence of quantisation. Analogue designs both deliver continuous control signals to the oscillators and have a continuous phase detection capability, unlike a digital system where these actions are carried out with fixed resolution.
\begin{figure}[!h]
	\centering
	\includegraphics[scale=0.4]{../ccirc_2013_arch}
	\caption[Architecture of the ADPLL network and of a single node]{Architecture of the ADPLL network and of a single node \cite{zianbetov2013distributed}.}
	\label{fig:eldar_node}
\end{figure}
Looking at the design of a given node it is notable that the function carried out by the Error Combiner is akin to a average, therefore as mentioned by Pratt and Nyugen, there is potential to a mode locked equilibrium in which the oscillators are not synchronised. In their paper, they presented a method where initial start-up was performed uni-directionally and, once all nodes are close to alignment, full connectivity could be restored, however this was not viable in an analogue system as reconfigurability was not an option \cite{pratt1995distributed}. Javidan \textit{et al} (2011) found that this was in fact the case, and stated:
\begin{quote}
	[Mode-locking] is solved in a simple and original way, by a dynamic reconfiguration of the network interconnection topology at the starting stage.
\end{quote}
In creating an entirely digital system, Zianbetov and Shan could easily reconfigure the network topology and by implement a uni-directional start-up avoid the problem of convergence into a mode locked state, without having to design a non-linear phase detector.

\section{\acs{ADPLL} Architecture}
As indicated in Figure \ref{fig:mulkeen_pll} the three main building blocks of a conventional \ac{PLL} are the \ac{PD}, \ac{LF} and \ac{VCO}. In an \ac{ADPLL} these blocks are then replaced by their digital counterparts, necessitating quantisation in order to remain physically realisable. The ``All-Digital'' moniker is a misnomer as the oscillator and \acl{PD} are usually both implemented by mixed signal circuits.
\begin{figure}[h]
	\centering
	\includegraphics[scale=0.5]{../tex_files/mulkeen_pll}
	\caption[Block fiagram of a \acl{PLL}]{Block diagram of a \acl{PLL}, \textit{Wireless Systems Notes}, B. Mulkeen (2017).}
	\label{fig:mulkeen_pll}
\end{figure}

\subsection{Digitally Controlled Oscillator}
In a digital system there are a very limited number of voltages representable, most commonly just two, so using a voltage to control the oscillator is not a viable strategy. Instead a fixed bit width signal is used to control the oscillator's period, selecting the number of inverters in a ring oscillator or the varactor configuration of a travelling wave oscillator \cite{chen2011rotary}. The decisions made in the design of the \ac{DCO}, or \ac{NCO}, determine many of the other \ac{ADPLL} parameters. While tuning range and centre frequency, as well as linearity, carry over from the analogue counterpart, a \ac{DCO} also has a frequency step which in combination with the bit width of the control signal determines the range over which the oscillator can be tuned. Figure \ref{fig:my_ring} illustrates a basic ring oscillator design. A ring oscillator is an inherently unstable circuit composed of an odd number of inverters connected in a circle, which allows a signal to propagate infinitely, with the signal at any point in the circuit appearing as a square wave. The half-period of this oscillator is the time taken for the signal to propagate once through the chain, $n$ times the propagation delay through one inverter. The frequency of operation can then be set by modulating the length of the chain, in steps of two inverters to maintain an odd number, by means of \texttt{f\_select}. The main impact of output frequency quantisation is that only frequencies which are integer multiples of the frequency step away from the centre frequency can be easily reproduced, with intermediate values only obtainable in a manner akin to Fractional-N synthesis with the control code toggling back and forth. This acts as a source of jitter in the system.
\begin{figure}[h]
	\centering
	\includegraphics[width=0.6\textwidth]{../inverter_chain}
	\caption[Basic Ring/Inverter Chain Oscillator]{Basic ring/inverter chain oscillator.}
	\label{fig:my_ring}
\end{figure}

@@

\begin{figure}[h]
	\centering
	\includegraphics[width=0.6\textwidth]{../shan_bb_pd}
	\caption[Bang-bang phase/frequency detector architecture]{Bang-bang phase/frequency detector architecture \cite{shan2014phd}.}
	\label{fig:shan_bb_pd}
\end{figure}

\subsection{Digital Loop Filter}
\begin{figure}[h]
	\centering
	\includegraphics[width=0.6\textwidth]{../simple_pi}
	\caption[Basic \ac{PI} controller architecture]{Basic \ac{PI} controller architecture.}
	\label{fig:my_simple_pi}
\end{figure}
The \acl{LF} in an \ac{ADPLL} can be implemented as a \ac{PI} controller, as only a low-pass filter is required, such as that in Figure \ref{fig:my_simple_pi}. In the case of a node in an \ac{ADPLL} network the input of this filter is a weighted of the phase difference relative to the neighbouring local clocking areas. In one example topology a digital system the proportional section can be implemented by a simple multiplier, whereas the integral path is constructed by adding the result of a multiplication by the proportional gain to an accumulator. This delayed summation can be easily implemented by an accumulator to which the current value of the multiplication is added each cycle, and as such the system has an infinite impulse response. The value of these gains determine the response and stability of the \ac{ADPLL} network. The transfer function of such a controller is given by \cite{shan2014phd}:
\begin{equation*}
	H(z) = \alpha + \beta\frac{1}{1-z^{-1}} = \frac{(\alpha + \beta) - \alpha z^{-1}}{1-z^{-1}}\vspace{-0.2cm}
\end{equation*}
It has been found by Koskin \textit{et al} (2018) that stable operation can be achieved when the integral gain, $k_i$, is less than half the proportional gain, $k_p$ \cite{koskin2018generation}. In the same study a range of values was found which would produce low jitter operation of the network. As these values are all less than one, the filter must implement fixed point arithmetic in an effort to maintain the simplicity of the clock distribution network, rather than incurring the penalty of floating point calculations.

@@

A similar technique was used by Lata \textit{et al} (2013) to implement a \ac{FPGA} clocked oscillator with a $200~\si{\kilo\hertz}$ centre frequency, however it is hard to understand exactly what either the goal was or the specifics of the \ac{DCO} used \cite{lata2013adpll}.

@@

\ac{FPGA}s have been used in other fields to simulate and experiment with new technologies, although not all of these attempt to implement the mixed-signal circuitry using primitive elements. Fernandez-Alvarez \textit{et al} (2016) proposed a method for prototyping high-voltage system controllers suited to higher end \ac{FPGA}s in which a co-processor on the \ac{FPGA} simulates the mixed-signal circuitry while the digital section of the design is implemented on the \ac{FPGA} itself \cite{fernandez2017hw}. While the interfacing between hardware and software remains a challenge they found:
\begin{quotation}
	Obtained data are compared to the data obtained by means of using ... simulation. The proposed solution speeds up the evaluation in around one order of magnitude keeping the accuracy. The output signal differs in less than 0.6 mV (RMSD).
\end{quotation}

Mixed-signal circuits were, however, simulated in hardware by \'{O}scar Luc\'{i}a \textit{et al} (2011) on an \ac{FPGA} in order to overcome the excessive time penalty imposed by software based simulations that required the behaviour of both a digital and mixed-signal peripheral and that of a micro controller running code in C to be simulated side by side \cite{lucia2011real}. They concluded
\begin{quotation}
	... the proposed system provides a versatile and fast method to develop ad hoc control architectures, avoiding the need for time-consuming mixed-signal simulations and the risk of damaging the actual power converter implementation.
\end{quotation}
By carrying out simulations on an \ac{FPGA}, Guanhua Wang \textit{et al} (2013) achieved a 3000 times decrease in run-time when compared to an identical simulation in MATLAB used for the verification of a calibration algorithm for successive approximation analogue-to-digital converters \cite{wang2013fast}. Many other examples can be found of \ac{FPGA}s used for the simulation of mixed-signal or \acs{RF} circuits.
%TODO Elena had suggestions?

@@

Both above forms of measurement assume a large number of sequential measurements, however, there are other ways that jitter and skew can be calculated \cite{AN10007}. Phase jitter is an important characteristic for communications systems as it can be used to calculate phase noise, important to ensure spurious emissions are within regulations. Long term, or accumulated, jitter represents the cumulative effect of jitter on the signal over several cycles. Long term jitter in particular affects \acs*{RADAR} as it will manifest itself as a Doppler shift in the return signal. For an \ac{PLL} network, with an appropriately designed filter, the long term jitter should be zero so long as the reference signal is stable. %TODO ask Brian about this one

### 2019-04-20 thesis/Conclusion.tex
? pages
79git statusgit statusgit statusgit statusgit statusgit statusgit statusgit statusgit statusgit statusgit statusgit status

\section{Conclusions}
\section{Impact of Research}
\section{Suggested Future Work}

>>>>

\section{Conclusions}
\section{Suggested Future Work}
\subsection{\acl{TDL} Characterisation}
\acreset{TDL}
As previously mentioned, characterisation of the \ac{TDL} was not carried out due to the time it would have taken to perform the required measurements and tests. This characteristion can be accomplished through the use of an off-chip tunable delay. Off-chip is unfortunately a requirement, as the only delays available of the order required to test the step sizes are the variable delay of the inverters themselves. A resistor \& capacitor delay circuit can be used to perform these test, and the output of the \ac{PFD} analysed on a mixed-signal oscilloscope alongside the delayed and reference waveforms. The use of an off-chip delay element necessitates careful construction/cailbration of the measurement environment to eliminate the delay stemming from the signal path through input/output buffers as it leaves the chip.

\subsection{Increased Network Size}
A 4x4 network was a stretch-goal of this project, depending on the time remaining after once the minumim goal of a 3x3 network had been reached. Unfortuantely due to problems with the clocked phase detector which resulted in mode-locking, incorrect \ac{RO} inverter selection logic and the projected time requirement to ensure alignment of each \ac{RO} this was not achieved. A 4x4 network would better explain some dynamics, such as the skew growth as the \ac{ADPLL} indices increase. However. progressing beyond a 3x3 network means that less than $20\%$ of \ac{ADPLL} output signals can be examined at any given point in time. As such, it may be useful to develop a more advanced measurement set-up that can either capture more signals simultaneously, or implement a measurement system that makes more efficient use of the available probes.

\subsection{\acs{FPGA} Clocked, Linear Period \acs{DCO}}
\acreset{FPGA,DCO}
While this design of \ac{DCO} was proposed earlier in this thesis, it was not implemented in any \ac{ADPLL}s, as it was only realised that the output waveform did not matter, just the location of the rising edges was not made until later in the project, once the pitfall encountered in the clocked \ac{PFD} design was resolved. Until this point the non-square wave output had eliminated the design from consideration. However, as in an \ac{ASIC} linear period designs are easier to implement, it would be ideal to have the benefit of frequency scaled hardware simulations with a \ac{DCO} that more closely relates the the eventual system.

\subsection{Procedural Network Instantiation}
This suggestion is directed at making the user experience simpler. As implemented, networks based on \ac{ADPLL}s containing \ac{RO}s are implemented in separate files to their \ac{FPGA} clocked \ac{DCO} based counterparts as are the networks of difference sizes. At a network size of 2x2 this was not a problem, but at the scale of 3x3 network the modules became somewhat cumbersome to work with, with a large number of interconnects. For 3x3 and larger networks it would be expedient to create the network with a \texttt{generate} statement, as this would reduce the number of lines of code and the amount of named interconnects required. Instead the user could set a number of configuration parameters and the network would be created accordingly. This could even allow for the use of different \ac{ADPLL} types in the same network with ease. Unfortunately the tasks of floorplanning and oscillator alignment would still be left to the user to perform.

%TODO cahracterise tdl

### 2019-04-20 thesis/FPGABasedADPLLs.tex
79git statusgit statusgit statusgit statusgit statusgit statusgit statusgit statusgit statusgit statusgit statusgit status

In being \ac{FPGA} clocked this design has pseudo-deterministic characteristics, with each period step being almost identical across oscillators and over the entire tuning range, unlike an \ac{ASIC} where process variation will impact the layout of a high frequency oscillator. The only variation in this design will come, ironically, from jitter or skew in the \ac{FPGA}'s clock distribution network, which as the frequencies will typically be in the low hundreds of MHz is very minor. In the case of the Xilinx \acl{Nexys} this is at most 100 picoseconds, or 0.05\% of the period of an intended output clock at 5 MHz. To put this value into perspective, on this board the minimum value of $T_{FPGA}$ that could be used to drive this oscillator is 3.87 nanoseconds, 1.935\% of the period.

@@

\subsection{\ac{FIR} \acl{LF}}
The \ac{FIR} implementation of the \acl{LF} will take the form presented in Figure \ref{fig:fir_pi} (a), and consists series of time delays, or ``taps'', the output of which is multiplied by a coefficient and the result of each multiplier added together to form the filter output. Various tools, Matlab for example, can be used to compute the coefficients and number of taps required to obtain a specific filter response. Using an \ac{FIR} filter is advantageous as it is always stable due to the lack of a feedback path, however a large number of taps are required to obtain a frequency response akin to that of the \ac{IIR} implementation, thus requiring an equal number of multipliers and $taps-1$ adders. Compared to an \ac{IIR}, which requires just a two adders, two multipliers and a flip-flop this has a much greater hardware footprint. It is possible to condense the \ac{FIR} multiplication and addition to a single multiplier and adder, depicted in \ref{fig:fir_pi} (b), however this will take as many clock cycles to complete as there are taps. This, however, is not viable in this system as the filter must produce its output in a single clock cycle. In the case of any coefficient symmetry, addition can be performed prior to multiplication, thus reducing the number of multipliers for each identical coefficient.
\begin{figure}[h]
	\centering
	\subfloat[Typical FIR Filter]{{\includegraphics[width=0.6\textwidth]{../FIR_1} }}\\
	\subfloat[Single Multiplier FIR Filter]{{\includegraphics[width=0.6\textwidth]{../FIR_2} }}%
	\caption[\ac{FIR} \ac{PI} loop filter RTL diagrams]{\ac{FIR} \ac{PI} loop filter RTL diagrams.}
	\label{fig:fir_pi}
\end{figure}

@@

\subsection{\ac{IIR} \acl{LF}}
A \ac{PI} controller can be easily implemented as an \ac{IIR} filter in hardware, as in Figure \ref{fig:iir_pi}, requiring significantly fewer circuit elements than an \ac{FIR} based design as the long series of taps and multipliers is not required, as instead integration is performed by a single multiplier and an accumulator. As before the frequency response of the controller can be calculated using the following equation, where $\alpha$ is the proportional and $\beta$ the integral gain:
\begin{equation*}
	H(z) = \alpha + \beta\frac{1}{1-z^{-1}} = \frac{(\alpha + \beta) - \alpha z^{-1}}{1-z^{-1}}
\end{equation*}

>>>>

In being \ac{FPGA} clocked this design has pseudo-deterministic characteristics, with each period step being almost identical across oscillators and over the entire tuning range, unlike an \ac{ASIC} where process variation will impact the layout of a high frequency oscillator. The only variation in this design will come, ironically, from jitter or skew in the \ac{FPGA}'s clock distribution network, which as the frequencies will typically be in the low hundreds of MHz is very minor. In the case of the Xilinx \acl*{Nexys} this is at most 100 picoseconds, or 0.05\% of the period of an intended output clock at 5 MHz. To put this value into perspective, on this board the minimum value of $T_{FPGA}$ that could be used to drive this oscillator is 3.87 nanoseconds, 1.935\% of the period.

@@

\subsection{\acs{FIR} \acl{LF}}
The \acl{FIR} implementation of the \acl{LF} will take the form presented in Figure \ref{fig:fir_pi} (a), and consists series of time delays, or ``taps'', the output of which is multiplied by a coefficient and the result of each multiplier added together to form the filter output. Various tools, Matlab for example, can be used to compute the coefficients and number of taps required to obtain a specific filter response. Using an \ac{FIR} filter is advantageous as it is always stable due to the lack of a feedback path, however a large number of taps are required to obtain a frequency response akin to that of the \ac{IIR} implementation, thus requiring an equal number of multipliers and $taps-1$ adders. Compared to an \ac{IIR}, which requires just a two adders, two multipliers and a flip-flop this has a much greater hardware footprint. It is possible to condense the \ac{FIR} multiplication and addition to a single multiplier and adder, depicted in \ref{fig:fir_pi} (b), however this will take as many clock cycles to complete as there are taps. This, however, is not viable in this system as the filter must produce its output in a single clock cycle. In the case of any coefficient symmetry, addition can be performed prior to multiplication, thus reducing the number of multipliers for each identical coefficient.
\begin{figure}[h]
	\centering
	\subfloat[Typical FIR Filter]{{\includegraphics[width=0.6\textwidth]{../FIR_1} }}\\
	\subfloat[Single Multiplier FIR Filter]{{\includegraphics[width=0.6\textwidth]{../FIR_2} }}%
	\caption[\ac{FIR} \ac{PI} loop filter RTL diagrams]{\ac{FIR} \ac{PI} loop filter RTL diagrams.}
	\label{fig:fir_pi}
\end{figure}

@@

\subsection{\acs{IIR} \acl{LF}}
A \ac{PI} controller can be easily implemented as an \acl{IIR} filter in hardware, as in Figure \ref{fig:iir_pi}, requiring significantly fewer circuit elements than an \ac{FIR} based design as the long series of taps and multipliers is not required, as instead integration is performed by a single multiplier and an accumulator. As before the frequency response of the controller can be calculated using the following equation, where $\alpha$ is the proportional and $\beta$ the integral gain:
\begin{equation*}
	H(z) = \alpha + \beta\frac{1}{1-z^{-1}} = \frac{(\alpha + \beta) - \alpha z^{-1}}{1-z^{-1}}
\end{equation*}
